Unfolding the genome's 3D architecture:\\ \Large{Insights in {\em P. falciparum} genome organization}

### 2017-07-27 doc/manuscript/paper.tex
First draft of the abstract

%\documentclass[2columns]{article}

\documentclass[letterpaper,12pt]{article}
\usepackage{url}
\usepackage{graphicx}
\usepackage{url}
\usepackage{wrapfig}
\usepackage[normalem]{ulem}
\usepackage{float}
\usepackage{xr}

@@

%% Load De Gruyter specific settings 
\usepackage{dgjournal}

@@

% Title of paper
\title{The art of modelling the 3D structure of the genome\\
  Insights in {\em P. falciparum} genome organization}
\author{Nelle Varoquaux}

%\maketitle

\begin{abstract}

@@

While efficacious vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citep{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms. Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at
the epigenetic and post-translational levels \citep{cui:chromatin-mediated,
duffy:role, hoeijmakers:placing, horrocks:control, deitsch:mechanisms}:
a better understanding of genome architecture of {\em P. falciparum}, both at
a local and global scale may provide new leads for developing new therapies.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{schemas/schema.pdf}
\caption{The art of understanding 3D genome structure}{
\fixme{write caption}}
\label{Fig1}
\end{figure}

@@

The development of genowe-wide and high-throughput protocols to probe samples
for their 3D genome architecture naturally paved the way to systematic and in
depth study of the folding mechanisms of DNA. Over the past 10 years, the
challenge of building 3D models from contact maps have been tackled by a
plethora of methods, some developped for a particular organisms, others with
the hope to be generalizable to any data sets. Methods model chromosomes as a
series of beads, and attempt to place the beads in a 3D euclidean space to
well represent the contact map. ``data-driven'' methods broadly fall into two
categories: (i) \textit{consensus} approaches, that aim at inferring a unique
mean structures best representing the contact count data; (ii)
\textit{ensemble} methods that yield a population of structures.

@@

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{figures/counts_maps.png}
\caption{Contact maps of {\em P. falciparum}'s chr7}{
\fixme{write caption}
\textbf{A.} \citep{lemieux:genome-wide}}
\label{Fig1}
\end{figure}

@@

A crucial step of MDS-based method is the conversion of counts into
wish-distances. As described earlier, converting contact counts into
wish-distances require strong assumptions that may not be met in practice. For
example, this mapping changes from one organism to another
\citep{fudenberg:higher-order}, from one resolution to another
\citep{zhang:inference} or even from one time point to another during the cell
cycle \citep{le:high-resolution, ay:three-dimensional}. To alleviate this
problem, after filtering interaction counts based on significance and
interpolation of missing values, \citet{ben-elazar:spatial} casts the
inference as a \emph{non-metric MDS} \citep{kruskal:multidimensional}, where
the 3D structure is inferred jointly with the wish-distances. Another idea is
to parametrize the count-to-distance mapping as a power-law ($d = \beta
c^\alpha)$, and to infer jointly the parameters jointly with the structure
\citet{zhang:inference}.

@@

\texttt{Pastis} can automatically adjust the parameters $\beta$, $\alpha$ of
the counts-to-distance transfer function and infer a genome structure that
best explains the observed data. The strength of this methods comes from the
robustness to low signal-to-noise ratio, through a direct modeling of the
noise through the statistical model.

\fixme{add results on the P. falciparum to show pastis's results on 
super noisy data}

\subsection*{Ensemble methods}

@@

\citet{umbarger:three-dimensional,bau:three-dimensional, kalhor:genome} models
chromosomes as a series of beads, linked by restraining oscillators. These
oscillators can be thought of as a ``force'' between beads so that they come
into contact or ensure a minimal or maximal distance between those. The models
includes three differente types of restraints: (i) beads seen interacting are
restrained with harmonic oscillators of strengths derived from the contact
counts; (ii) adjacent beads are ensured to be neither too close nor too far
from one another; (iii) the last oscillators are activated only when
constraints are not fulfilled to XXX.  This yields an optimization problem
with a large number of local minima, which the authors sample from by running
50,000 minimizations starting from random starts.

@@

\cite{tjong:physical, tokuda:dynamical, tjong:physical} model {\em S.
cerevisiae}'s chromosomes as flexible random fibers under a small set of
constraints. While the exact modeling proposed by the three groups differ, the
set of constraints can roughly be summarized as: \textit{(i)} the chromosomes
are constrained into a spherical ball representing the nucleus; \textit{(ii)}
centromeres are constrained into a spherical ball tethered to the nuclear
membrane; \textit{(iii)} telomeres are tethered to the nuclear membrane;
\textit{(iv)} rDNA is constrained into the nucleolus, represented as a
spherical ball opposite to the centromeres. A scoring function eng

Using IMP, the authors generate a 25,000 structures, all of which respect the
constraints listed above. The population of structures is then used to created
a ``volume-exclusion contact map'', considering that two beads that are less
then 45~nm apart from a contact. The Pearson correlation of the
volume-exclusion contact map and the Hi-C one are highly correlated,
demonstrating this small set of constraints fully explain the observed counts.
In addition, the population of structures also explain FISH experiments
previously published.

@@

The second question one can ask is how do the structures differ. Tackling this
question is very challenging, but can be reformulated in a much easier way.
Are the hallmarks of interest conserved across structures of a same stage?
\citet{ay:three-dimensional} and \citet{lemieux:genome-wide} both identified
{\em P. falciparum} folded in very specific ways, with VRSSM genes highly
interacting. \citet{ay:three-dimensional} also observed strong clustering of
the centromeres, and enrichment in interaction at the telomeres. These
observation can lead to a rigorous approach at identifying whether families
of loci clustered in the structures and even quantifying this in a rigourous
manner.

\subsection*{Chromatin compaction}

Bau et al

@@

data-driven Plethora of methods, but only handfull that can be used on different
organisms and have been validated on simulated data set. Ensemble methods are
more challenging to use, to validate and to explore, but offer more
biologically accurate view of what is going on in teh nucleus. Overall,
powerfull tool but that still needs to be extended oto work on challenging
data set.

model-driven methods are organisms and study specific. For each data set, the
set of constraints need to be revisited. Yet, once built, they can be used ot
replace high throughput FISH experiments at low cost.

Downstream analysis gave important insights in the relation between gene
regulation and genome architecture of teh {\em P. falciparum}. Yet a lot
remains to be done. How to compare in a meaningful manner structures from
different time points at a genome-scale level?

>>>>

\documentclass[2columns]{article}

%\documentclass[letterpaper,12pt]{article}
\usepackage{url}
\usepackage{graphicx}
\usepackage{url}
\usepackage{wrapfig}
\usepackage[normalem]{ulem}
\usepackage{float}
\usepackage{xr}

@@

%% Load De Gruyter specific settings 
%\usepackage{dgjournal}

@@

% Title of paper
\title{Unfolding the genome's 3D architecture:\\
  \Large{Insights in {\em P. falciparum} genome organization}}
\author{Nelle Varoquaux}

\maketitle

\begin{abstract}
\begin{itemize}

\item {\em P. falciparum} is the deadliest parasite causing human malaria,
responsible for more than 500,000 death in 2015. Our lack of understanding of
the parasite gene regulation of its complex life cycle is an impediment to the
development of novel therapies. Recent evidence points towards unusual means
of gene regulation, possible through changes in its genome architecture. A
better understanding of how DNA folds into the nucleus and how it influences
gene expression could provide new leads for developing new drugs.
\item In the meantime, the development of new means to probe samples for the
three-dimensional structure of DNA paves the way for in depth and systematic analysis of genome
architecture, offering new insights in many biological processes. 3C-like
methods can now assess physical interactions between pairs of loci in a
genome-wide fashion, thus leading to the creation of contact maps. The
democratization of such protocols creates many new opportunities for
methodological development: how to infer 3D models from these contact maps?
Can such models help us gain insights in biological process?
\item Here, I review recent methodological developments \fixme{finish this} 
\end{itemize}

@@

While safe and effictive vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citep{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms. Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at
the epigenetic and post-translational levels \citep{cui:chromatin-mediated,
duffy:role, hoeijmakers:placing, horrocks:control, deitsch:mechanisms}:
a better understanding of genome architecture of {\em P. falciparum}, both at
a local and global scale may provide new leads for developing new therapies.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{schemas/schema.pdf}
\caption{The art of understanding 3D genome structure}
\label{Fig1}
\end{figure}

@@

The development of genome-wide and high-throughput protocols to probe samples
for their 3D genome architecture naturally paved the way to systematic and in
depth study of the folding mechanisms of DNA. Over the past 10 years, the
challenge of building 3D models from contact maps have been tackled by a
plethora of methods, some developed for a particular organisms, others with
the hope to be generalizable to any data sets. Methods model chromosomes as a
series of beads, and attempt to place the beads in a 3D euclidean space to
well represent the contact map. ``data-driven'' methods broadly fall into two
categories: (i) \textit{consensus} approaches, that aim at inferring a unique
mean structures best representing the contact count data; (ii)
\textit{ensemble} methods that yield a population of structures.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.7\linewidth]{figures/counts_maps.png}
\end{center}
\caption{Contact maps of {\em P. falciparum}'s chr7}{ \small Both contact maps show
enrichment of contact counts in VRSSM clusters (\textbf{A.} Ring stage 
\citep{lemieux:genome-wide} of IT strains, \textbf{B.} 
Trophozoite stage of 3D7 \citep{ay:three-dimensional}) 
Dark blue corresponds to high interactions, light yellow to low interactions.
The strong diagonal reflects the proximity of adjacent regions of the genome.
The black line indicates the centromeres. }
\end{figure}

@@

A crucial step of MDS-based method is the conversion of counts into
wish-distances. As described earlier, converting contact counts into
wish-distances require strong assumptions that may not be met in practice. For
example, this mapping changes from one organism to another
\citep{fudenberg:higher-order}, from one resolution to another
\citep{zhang:inference} or even from one time point to another during the cell
cycle \citep{le:high-resolution, ay:three-dimensional}. To alleviate this
problem, after filtering interaction counts based on significance and
interpolation of missing values, \citet{ben-elazar:spatial} casts the
inference as a \emph{non-metric MDS} \citep{kruskal:multidimensional}, where
the 3D structure is inferred jointly with the wish-distances. Another idea is
to parametrize the count-to-distance mapping as a power-law ($d = \beta
c^\alpha)$, and to infer jointly the parameters jointly with the structure
\citep{zhang:inference}.

@@

\texttt{Pastis} can automatically adjust the parameters $\beta$, $\alpha$ of
the counts-to-distance transfer function and infer a genome structure that
best explains the observed data. The strength of this methods comes from the
robustness to low signal-to-noise ratio, through a direct modeling of the
noise through the statistical model (See figure~\ref{fig:3D_structures}.

\begin{figure}
\includegraphics[width=\linewidth]{schemas/3D_structures.pdf}
\label{fig:3D_structures}
\caption{3D models of {\em P. falciparum}'s genome architecture}{Using
\texttt{pastis-MDS} (\textbf{A} and \textbf{B}) and \texttt{pastis-PO}
(\textbf{C} and \textbf{D}) on two
HiC data set of {\em P. falciparum}. Each color corresponds to a chromosome.
Large white beads mark centromeres, small blue beads telomeres and green beads
VRSSM clusters. Centromeric cluster is marked with a black line, telomeric
cluster is marked with a dashed line.}
\end{figure}

\subsection*{Ensemble methods as a mean to infer population of structures}

@@

\citet{umbarger:three-dimensional,bau:three-dimensional, kalhor:genome} models
chromosomes as a series of beads, linked by restraining oscillators. These
oscillators can be thought of as a ``force'' between beads so that they come
into contact or ensure a minimal or maximal distance between those. The models
includes three different types of restraints: (i) beads seen interacting are
restrained with harmonic oscillators of strengths derived from the contact
counts; (ii) adjacent beads are ensured to be neither too close nor too far
from one another; (iii) the last oscillators are activated only when
constraints are not fulfilled to \todo{XXX}.  This yields an optimization problem
with a large number of local minima, which the authors sample from by running
50,000 minimizations starting from random starts.

@@

\cite{tjong:physical, tokuda:dynamical, tjong:physical} model {\em S.
cerevisiae}'s chromosomes as flexible random fibers under a small set of
constraints. While the exact modeling proposed by the three groups differ, the
set of constraints can roughly be summarized as: \textit{(i)} the chromosomes
are constrained into a spherical ball representing the nucleus; \textit{(ii)}
centromeres are constrained into a spherical ball tethered to the nuclear
membrane; \textit{(iii)} telomeres are tethered to the nuclear membrane;
\textit{(iv)} rDNA is constrained into the nucleolus, represented as a
spherical ball opposite to the centromeres. One can then simulate a large set
of random structures fulfilling the constraints, and generate a
``volume-exclusion contact map'' from this population of structures,
considering that two beads that are less then 45~nm apart in any of the
structures form a contact. The Pearson correlation of the volume-exclusion
contact map and the Hi-C one are highly correlated, demonstrating this small
set of constraints fully explain the observed counts. In addition, the
population of structures also explain FISH experiments previously published.

\subsection*{Volume-exclusion modeling cannot explain {\em P. falciparum}'s
genome architecture}

For at least some stages, {\em P. falciparum} share a lot of common genomic
architecture hallmarks with the budding yeast: the centromeres are strongly
co-localized at one end of the nucleus, telomeres are in physical contacts,
\dots So could the primary architectural features also arise from a population
of constrained but otherwise random of structures? Adapting the set of
constraints to match biological knowledge of {\em P. falciparum},
\citet{ay:three-dimensional} showed the resulting simulated VE map not only
yielded lower correlations, but also did not show the same features as the
original contact count matrix. In particular, the VRSM genes which display
domain-like enrichment in interactions did not appear in the simulated VE map
(See Figure~\ref{fig:ve_modeling})

\begin{figure}
\caption{Volume Exclusion Modeling}{}
\label{fig_ve_modeling}
\end{figure}

@@

The second question is: ``are models locally consistent''
\citep{bau:three-dimensional}. A possible answer to this questions is to
divide the structures into overlapping sub-structures ranging from 5 to 20
beads, and compute the pairwise RMSD across all structures. Segments
overlapping within a certain range for a large number of models can be
assessed as locally consistent, while others labelled as highly variable.

The last question one can ask is: ``how do the structures differ?'' . Tackling
this question is very challenging, but can be reformulated in a much easier
way. Are the hallmarks of interest conserved across structures of a same
stage? \citet{ay:three-dimensional} and \citet{lemieux:genome-wide} both
identified {\em P. falciparum} folded in very specific ways, with VRSSM genes
highly interacting. \citet{ay:three-dimensional} also observed strong
clustering of the centromeres, and enrichment in interaction at the telomeres.
These observation can lead to a rigorous approach at identifying whether
families of loci clustered in the structures and even quantifying this in a
rigourous manner.

\subsection*{Chromatin compaction and chromosome entanglement}

3D models can be used to estimated chromatin compaction and chromosome
entanglement. The former can be estimated by looking at the number of base
pairs in a region defined either by volume if the scale of the structure is
known, or by a percentage of the size of the structure
\citep{bau:three-dimensional}. Another idea is to sample random beads of a
certain diameter, and assess how many loci are seen interacting. Applying this
method on the 3 models of {\em P. falciparum}, \citet{ay:three-dimensional}
showed the trophozoite stage exhibited a more open chromatin than the ring and
schizont stage, coherent with the transcriptionally active state of the
parazite during this moment of the life cycle. A similar analysis, but
counting the number of inter-chromosomal interaction can help assess the
chromosome entanglement of the structure.

@@

Plethora of methods for inferring 3D models from contact maps have been
developed: some are very specific to an organism or a region of the genome,
while others are generalizable easily to many organisms. As 3C-type method are
democratized, a wider audience of researchers are in need of robust and
well-implemented algorithms for building models. Surprisingly, only a handfull
of methods are both publicly available, generalizable and well-validated.
While \textit{ensemble} methods offer a more biologically accurate view of the
variety of chromatin folding in a population of cells, their use is more
challenging, both as a result of a lack of comparison and validation of the
methods, but also as large population of structures are less amenable to
exploration and visualization than \textit{consensus} structures.

At the other end of the spectrum, model-driven methods are organisms and study
specific, and require good understanding of both polymer physics and of the
particular organism studied: for each data set, the set of constraints to
apply on the structure need to be revisited and remained. Yet, once built,
they provide extensive insights in the average location of genes and may be a
mean to replace high throughput FISH experiments at low cost as a first
exploration tool.

While downstream analysis of the {\em P. falciparum} gave important insights
in the relation between gene regulation and genome architecture of the {\em P.
falciparum}, the use of 3D models for downstream analysis remains scarce in
the literature and many avenues are left opened to methodological development.
How to compare in a meaningful manner structures from different time points at
a genome-scale level?
